Numerical analysis of the linear vertical shear instability in astrophysical disks

### 2014-12-29 appendix.tex
restructing analytic discussion

\section{Full governing equation for vertically isothermal disks}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
 In \S\ref{analytic_adia} we made the replacement $D\to\Omega_k^2$
 before combining Eq. \ref{adia_iso1}---\ref{adia_iso2} to derive
 Eq. \ref{adia_iso3}.

This procedure ignores the vertical dependence of
 $D=\kappa^2(z) - \sigma^2$.

We show here that this has no significant
 consequence for thin disks.

@@

 Eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
 Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
 vertical dependence of $D$ and including thermal relaxation, we
 obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}

>>>>

\section{Full governing equation for vertically isothermal disks}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
eliminating variables to obtain a single equation for $\delta v_z$,
Eq. \ref{vertiso_gov}.  This procedure ignores the vertical dependence of
$D=\kappa^2(z) - \sigma^2$. We show here that this has no significant
consequence for thin disks.

@@

 Eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
 Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
 vertical dependence of $D$ and including thermal relaxation, we
 obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in
\S\ref{analytic_adia} and \S\ref{analytic_relax}.  This issue does not
arise for isothermal perturbations discussed in \S\ref{iso_discuss}
since in that case we work with a governing equation for $W$ instead.

### 2014-12-29 linear_adia.tex
restructing analytic discussion

\subsection{Perturbations with thermal relaxation}
Having examined the extreme limits of isothermal ($t_c\to0$) and
adiabatic ($t_c\to\infty$) perturbations, we now generalize further to consider
perturbations with intermediate thermodynamic responses with general
values of $t_c$ or $\beta$.

@@

We could proceed by considering the entire set of polynomial solutions
for $W$, and hence $\delta v_z$, discussed in \S\ref{iso_poly}, but it
is simplest to look at the fundamental mode. Thus we consider the `basic state' 
 \begin{align*}
   \delta v_z = 1,\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
   \ii\epsilon q \hat{k},
 \end{align*}
 then we may seek 
 \begin{align}
   \delta v_{z1} = a \zhat^2 + b,
 \end{align}
where $a$, $b$ are constants.

@@

\subsubsection{General dispersion relation with thermal relaxation}\label{disp_relax}
Eq. \ref{nearly_iso_explicit} is same form
as the governing equation for the adiabatic case
(Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
have a complicated dependence on $\hat{\sigma}$. 
Applying the same solution method as in \S\ref{adia_explicit}, we 
% Thus calculating the
% eigenfrequency $\hat{\sigma}$ generally requires a numerical
% approach, which we do  in \S\ref{disp_relax}. However, analytical
% results may be obtained in the limit $\beta\to0$. 
find the adiabatic dispersion relation,
Eq. \ref{adia_disp0}, generalizes to a $6^\mathrm{th}$ order
polynomial in $\hat{\sigma}$:
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
Eq. \ref{relax_disp} contrasts to Eq. \ref{adia_disp}, which is
quadratic in $\hat{\sigma}$ for adiabatic perturbations.

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $k_xH_\mathrm{iso}$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.

>>>>

\subsection{Perturbations with thermal relaxation}\label{analytic_relax}
Having examined the extreme limits of isothermal ($t_c\to0$) and
adiabatic ($t_c\to\infty$) perturbations, we now generalize further to consider
perturbations with intermediate thermodynamic responses with general
values of $t_c$ or $\beta$.

@@

We could proceed by considering the entire set of polynomial solutions
for $W$, and hence for $\delta v_z$, as derived in \S\ref{iso_poly}, but it
is simplest to look at the fundamental mode. Thus we consider the `basic state' 
 \begin{align*}
   \delta v_z = 1,\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
   \ii\epsilon q \hat{k},
 \end{align*}
 then we may seek 
 \begin{align}
   \delta v_{z1} = a \zhat^2 + b,
 \end{align}
where $a$, $b$ are constants.

@@

\subsubsection{Dispersion relation with thermal relaxation}\label{disp_relax}
Eq. \ref{nearly_iso_explicit} is same form
as the governing equation for the adiabatic case
(Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
have a complicated dependence on $\hat{\sigma}$. 
Applying the same solution method as in \S\ref{adia_explicit}, we 
% Thus calculating the
% eigenfrequency $\hat{\sigma}$ generally requires a numerical
% approach, which we do  in \S\ref{disp_relax}. However, analytical
% results may be obtained in the limit $\beta\to0$. 
find the adiabatic dispersion relation,
Eq. \ref{adia_disp0}, generalizes to a $6^\mathrm{th}$ order
polynomial in $\hat{\sigma}$ with finite thermal relaxation:
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
Eq. \ref{relax_disp} contrasts to Eq. \ref{adia_disp}, which is
quadratic in $\hat{\sigma}$ for adiabatic perturbations.

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $k_xH_\mathrm{iso}$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. Consistent with the previous
discussion, for $\beta\to0$ thermal relaxation has no effect as
$k_x\to0$ or $k_x\to\infty$.

Fig. \ref{relax_disp_fig} shows that a growing fundamental VSI mode
exists both in the isothermal ($\beta\to0$) and adiabatic
($\beta\to\infty$) limits, although the latter growth rates are
significantly smaller and perturbations are restricted to the
midplane. However, for $0.1 \lesssim \beta\lesssim
\beta 30$, this mode is stabilized for $k_xH_\mathrm{iso}\gtrsim
10$. Although results for small $k_x$ violates the low-frequency
approximation, Fig. \ref{relax_disp_fig} suggests that thermal
relaxation needs to be significantly shorter than the dynamical time
in order for the VSI with appreciable growth rates to develop.
